Data Provenance for Concurrent Program Debugging and Analysis in Production Systems

### 2013-11-09 paper.tex
committing many TODOs, a new fig, some new Background text and some copy/paste-blasted Eval text.  Lots to do!

\title{\Large Last Writer Slicing: Efficient Data Provenance for Concurrent Program Debugging and Analysis}

@@

\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.  
{\bf TODO Re-write}
We design efficient system support for collecting {\em last writer slices} for
shared-memory concurrent programs.  Last writer slices dynamically record the
code responsible for shared variable updates.  Last writer slices reflect the
provenance of values in memory, information that is useful during debugging.
Using last writer slices, we build {\em communication traps} (\ctraps).
\ctraps is a mechanism that efficiently monitors and can interpose on memory
operations that share data between threads.  \ctraps is an extensible
framework, exposing inter-thread dependences to {\em \ctraps applications} that
can implement arbitrary dependence analyses.

We implement last writer slicing and \ctraps.  We show that last writer slices
provide information that is useful for debugging several real-world bugs.  We
show \ctraps is useful by implementing two concurrency analyses from prior
work.  We evaluate our designs on a set of server programs and standard
benchmarks.  Our results show that collecting last writer slices has low run
time overheads (0--15\%) for many important applications, like memcached,
LevelDB, MySQL, and Apache.  We show that \ctraps imposes low run time
overheads by default and that the overheads scale with the complexity of the
analysis being implemented.

@@

In concurrent programs, collecting debugging information, like data provenance,
from production systems is particularly important.  Concurrency bugs manifest
as failures rarely.  When a deployed system fails and a programmer receives a
bug report, reproducing the reported failure to collect provenance information
may be difficult or impossible.  It is essential to collect as much information
as performance margins permit from deployed systems, to reduce the need for
time-consuming failure reproduction.

Prior work has shown that provenence helps with debugging~\cite{badapples,
whylineicse, tipslicingsurvey}.  One effort tracked code that wrote unusable
values into variables~\cite{badapples}.  That work showed that revealing the
connection between a bad value's origin and a failure it causes is helpful
during debugging.  Other work showed that programmers debug programs more
effectively when they can explicitly ask "why" questions about the origin of
certain values during debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea.  First, prior work~\cite{badapples} has focused on providing
provenance information for only some unusable {\em values}, like {\tt NULL}.
Limiting the applicability of their technique to only some failures, like {\tt
Null Pointer Exceptions}.  Second, while prior work has shown that limited
provenance tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples} impose overheads too high for production for C
and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Focusing on
concurrency is important because a value could have provenance in any of a
large number of threads, making reasoning about why a variable has a value
complex and difficult.

In that context, we design a novel form of in-production provenance tracking
for concurrent programs with three goals.   (1) To collect a general form of
provenance information, not limited to certain values.  (2) To provide
provenance information that helps understand how threads interact, which is
especially useful for concurrent program debugging.  (3) To work with overheads
low enough for use in production (ideally, less than 10\%).

@@

Data provenance information is directly useful during program debugging.  Work
on debugging~\cite{whylinechi,whylineicse} in the HCI and Software Engineering
community shows that when programmers debug programs, they benefit from the
ability to directly ask such provenance (``why'') questions of their debugger.
Note that that like prior work on debugging with provenance, in this work we
focus on {\em debugging} -- understanding the root cause of a failure -- rather than {\em bug detection} -- finding new, unknown bugs.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/BackgroundCoreDumpFail.pdf}
\caption{\label{fig:coreDumpFail}{\bf Ordering violation bug in HTTrack.} Simply examining memory does not provide provenance information that helps understand this bug.}
\end{figure}

Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
plays in debugging, using an example of a real bug from HTTrack, a web caching
tool.  The figure depicts two threads manipulating a data structure that
contains a mutex\_lock.  The program has an ordering violation bug: the
program crashes when Thread 2 tries to acquire the lock before Thread 1 has
initialized it.  Assuming the run time environment supports it, a core dump
delivered to the programmer after such a failure reveals very little about how
to fix the bug.  The lock is uninitizalized, so the core dump may show that it
holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
insight into the root cause of the failure, which is that the code permits a
use of the lock before its initialization.   Comparing the core dump from the
failing run to the memory image from a correct execution -- a helpful debugging
strategy -- reveals only that in a correct execution, the lock contains valid
mutex state.  This revalation is of limited use as it does not rule out
arbitrary data corruption as the bug's root cause.

Another challenge in debugging this bug is that it manifests as a failure
infrequently and under only real-world conditions.  If the programmer runs the
program in a debugger, the failure may not manifest in a reasonable amount of
time and running it may require unavailable real-world inputs.  Thus,
collecting provenance or other information offline is time-consuming or
impossible.  A better strategy is to continuously collect that information in
deployed systems and package it with core dumps sent to programmers with bug
reports.  In this work, we propose a last writer slicing design that espouses
this {\em in situ} data collection strategy and does so with overheads
appropriate for production systems.

%Dynamic dependence slices are especially useful for debugging if they are
%available after a failure has occurred in a deployed system. However, slicing
%in deployed systems is uncommon, because deployed systems demand low overheads
%and slicing can have high overhead.  Our system provides provenance information
%that is useful at debugging time by collecting {\em last writer slices} that
%help understand inter-thread communication.  We track last writer slices with
%overhead low enough for use in production code.  Deployable efficiency makes it
%possible to collect and report last writer slices the same way today's systems
%collect and report core dumps.

\subsection{Tracking Dependences and Communication}
\label{sec:background:comm}
%\ctraps tracks dependences that correspond to {\em inter-thread communication}.

Monitoring inter-thread dependences is the foundation of many useful
concurrency analyses.  For example, in a software transactional memory (STM)
system~\cite{stm} dependences correspond to interference between accesses in
different threads that must be detected to rollback conflicting transactions.
Deterministic concurrent runtime systems~\cite{coredet,grace} are similar --
they monitor inter-thread dependences, to determine when to apply
deterministic ordering constraints on threads' accesses.  STMs and
deterministic systems must track all dependences and they must track them
precisely, to provide strong guarantees as programming and execution models.
Unfortunately, precision has a high runtime overhead -- {\em e.g.},
CoreDet incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is
too high for use in production.

@@

@@

@@

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = \{q_{0}, \ldots, q_{e}\}$ where each $q \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:eval:debugging}
We illustrate the debugging benefits of last writer slices by using our system
to debug several real-world programs studied previously in the debugging
literature.

@@

\paragraph{HTTrack}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:debugging} describes the benefit provided by
last writer slices in debugging an ordering violation in HTTrack-3.43.9.

@@

Last writer slices help elucidate a subtle atomicity violation in the
StringBuffer library in version 1.4 of the Java JDK.  We use the C++ port of
the buggy JDK and a driver program code developed in prior concurrency debugging
work~\cite{concurrencybugs}.

Figure~\ref{fig:jdklws} shows how the bug can manifest as a failure. The string
object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
variable called {\tt count} that stores its length.  The {\tt append()}
code should atomically read {\tt sb->count} and update the string's buffer.
The failure occurs when the {\tt append()} code in Thread 1 is interleaved with
the {\tt erase()} code in Thread 2 as indicated in the figure.  The {\tt
erase()} code updates the buffer's length, making the version Thread 1 read
stale.  Thread 1 uses its stale version in the function called at point {\bf
B}, causing a crash.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug.pdf}
\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
\end{figure}

When the atomicity violation manifests, the appending thread is updating the
string's underlying buffer.  The erasing thread, however, may have completed
its erase operation.  The execution context of the erasing thread at the point
of the failure provides no insight as to why the program failed.  Examining the memory image, the programmer can only conclude that the length and buffer are 
inconsistent, but not why they are inconsistent.

This situation illustrates the value of the provenenace information furnished
by last writer slices.  At the point of the failure the last writer slice for
the length variable points to a decrement of that variable in the erase
function.  The last writer slice illustrates the need for synchronization
between the append function and the erase function.

>>>>

\title{Data Provenance for Concurrent Program Debugging and Analysis in Production Systems}

@@

\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.  
We propose {\em last writer slices}, an analysis for shared-memory
multi-threaded programs that tracks {\em data provenance information} in
production code.  Last writer slices dynamically record the thread and
operation that last wrote each variable, reflecting the {\em provenance} of
values.  Provenance helps understand bugs and we provide last writer slices to
programmers during debugging.  We build {\em communication traps}
(\ctraps) using last writer slices.  \ctraps efficiently
monitors and can interpose on operations that share data between
threads.  \ctraps is extensible, exposing inter-thread dependences
to {\em \ctraps applications} that implement arbitrary analyses.

We implement last writer slicing and \ctraps.  We conduct debugging studies of
real, buggy programs showing last writer slices are useful for debugging.  We
show \ctraps is useful by implementing two concurrency analyses from prior
work.  We evaluate the performance of our designs using server programs
and standard benchmarks.  We show that collecting last writer slices
has low run time overheads (0--15\%) for many applications,including  
memcached, LevelDB, MySQL, and Apache.  We show that \ctraps imposes low 
overheads by default and that overheads scale with analysis complexity.

@@

Collecting debugging information, like data provenance, from production systems
is particularly important for concurrent software.  Concurrency bugs manifest
as failures rarely.  When a production system fails and a programmer receives a
bug report, reproducing the reported failure to collect provenance information
may be difficult or impossible.  It is essential to collect as much information
as performance margins permit from production, to reduce the need for
time-consuming failure reproduction.

Prior work have previously shown that some limited forms of provenence help
with debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort
tracked code that wrote unusable values into variables~\cite{badapples}.  That
work showed that revealing the connection between a bad value's origin and a
failure it causes is helpful during debugging.  Other work showed that
programmers debug programs more effectively when they can explicitly ask "why"
questions about the origin of certain values during
debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea.  First, prior work~\cite{badapples} has focused on providing
provenance information for only some unusable {\em values}, like {\tt NULL}.
Limiting the applicability of their technique to only some failures, like {\tt
Null Pointer Exceptions}.  Second, while prior work has shown that limited
provenance tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples} impose overheads too high for production for C
and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Addressing
concurrency is important because a value could have provenance in any of a
large number of threads, making reasoning about why a variable contains that
value complex and difficult.

In this work we design a novel form of in-production provenance tracking for
concurrent programs with three goals.  (1) To collect a general form of
provenance information, not limited to only certain values.  (2) To provide
provenance information that helps understand how threads interact, which is the
key to understanding concurrency bugs.  (3) To work with overheads low enough
for use in production (ideally, less than 10\%).

@@

Data provenance information is directly useful during program debugging.  prior
work~\cite{whylinechi,whylineicse} community showed that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.  Origin tracking~\cite{badapples} and
program slicing~\cite{tipslicingsurvey} also make debugging easier by reporting
the provenance of values.  Like these prior efforts, in this work we focus on
using data provenance information for {\em debugging} -- understanding the root
cause of a failure -- as opposed to {\em bug detection} -- finding new, unknown
bugs.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferFail.pdf}
\caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
\end{figure}

Figure~\ref{fig:coreDumpFail} illustrates the role of provenance information in
debugging, using buggy code from the JDK-1.4 stringbuffer library, adapted to
C++ by other researchers and used variously in prior
work~\cite{concurrencybugs}.  The stringbuffer object pointed to by {\tt sb} in
Thread 1's code contains a buffer and a variable called {\tt count} that stores
its length.  The {\tt append()} code should atomically read {\tt sb->count} and
update the string's buffer (the full update code is not pictured).  The failure
occurs when the {\tt append()} code in Thread 1 is interleaved with the {\tt
erase()} code in Thread 2 as indicated in the figure.  The {\tt erase()} code
updates the buffer's {\tt count}, making the version that Thread 1 read into
{\tt len} stale.  Thread 1's assertion fails at {\bf B} using the stale {\tt
len} value, causing a crash.

When the failure manifests, Thread 1 is at the assertion in {\tt append()}, as
picture.  However, Thread 2 may have completed its erase operation and moved on
to execute some unrelated code.  The call stack for Thread 2 may show nothing
about its involvement in the failure, despite the fact that its interleaving
with the {\tt append()} code is the root cause.  Examining the core dump, the
programmer can see that the length and buffer are inconsistent, but is left to
ask {\em why} they are inconsistent.

Understanding the connection between {\tt append()} and {\tt erase()} is the
key to fixing this bug.  The provenance of the value stored in {\tt sb->count}
at {\bf B} was the write at {\bf A} in {\tt erase()} and with that provenance
information, the programmer would understand that key connection.  Prior
provenance work on origin tracking for unusable values~\cite{badapples} does
not help here.  The involved values are inconsistent, but they are all {\em
usable} so their provenance would not be tracked.

Another challenge posed by this bug is that it manifests as a failure
infrequently and under only real-world conditions.  If the programmer runs the
program in a debugger, the failure may not manifest in a reasonable amount of
time and running it may require unavailable real-world inputs.  Thus,
collecting provenance or other information offline may be time-consuming or
impossible.  A better strategy is to continuously collect that information in
deployed systems and package it with core dumps sent to programmers with bug
reports.  In this work, we propose a last writer slicing design that espouses
this {\em in situ} data collection strategy and does so with overheads
appropriate for production systems.

\subsection{Tracking Dependences and Communication}
\label{sec:background:comm}
{\bf TODO: Clean this up, downplay the "dependences" obsession and make it 
a better match to the LWS mission, as stated in the intro -- provenance + instrumentation = analysis}
Monitoring inter-thread dependences is the foundation of many useful
concurrency analyses.  For example, in a software transactional memory (STM)
system~\cite{stm} dependences correspond to interference between accesses in
different threads that must be detected to rollback conflicting transactions.
Deterministic concurrent runtime systems~\cite{coredet,grace} are similar --
they monitor inter-thread dependences, to determine when to apply
deterministic ordering constraints on threads' accesses.  STMs and
deterministic systems must track all dependences and they must track them
precisely, to provide strong guarantees as programming and execution models.
Unfortunately, precision has a high runtime overhead -- {\em e.g.},
CoreDet incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is
too high for use in production.

@@

{\bf TODO: Fix this figure to reflect availability of LWS info if the program crashes/is stopped at *any* point in the exec (like the read X in thread 2).}

@@

{\bf TODO: Replace with the "fixed" version of the JDK figure (currently in eval).  Update the text to match the new figure.}

@@

\subsection{\lwt Formalism and Correctness}
{\bf TODO: Necessary?  Is it really interesting?}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = \{q_{0}, \ldots, q_{e}\}$ where each $q \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:eval:debugging}

{\bf TODO: Table showing the bugs we looked at, vital stats, comparison to origin tracking}

We illustrate the debugging benefits of last writer slices by using our system
to debug several real-world programs studied previously in the debugging
literature.

@@

\paragraph{HTTrack}
{\bf TODO: Deal with this duplication from the copy/paste}
Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
plays in debugging, using an example of a real bug from HTTrack, a web caching
tool.  The figure depicts two threads manipulating a data structure that
contains a mutex\_lock.  The program has an ordering violation bug: the
program crashes when Thread 2 tries to acquire the lock before Thread 1 has
initialized it.  Assuming the run time environment supports it, a core dump
delivered to the programmer after such a failure reveals very little about how
to fix the bug.  The lock is uninitizalized, so the core dump may show that it
holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
insight into the root cause of the failure, which is that the code permits a
use of the lock before its initialization.   Comparing the core dump from the
failing run to the memory image from a correct execution -- a helpful debugging
strategy -- reveals only that in a correct execution, the lock contains valid
mutex state.  This revalation is of limited use as it does not rule out
arbitrary data corruption as the bug's root cause.

@@

The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:debugging} describes the benefit provided by
last writer slices in debugging an ordering violation in HTTrack-3.43.9.

{\bf TODO: Move this to Section 3 where it will live.}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug.pdf}
\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
\end{figure}
